Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-05-16 talk/dls2014/paper/paper.tex
adjust some details

If we start multiple such transactions in multiple threads, the TM
system guarantees that the outcome of running the transactions is
\emph{serialisable}. Meaning, the outcome is equal to some sequential
execution of these transactions. This means that the approach provides the same
semantics as using the GIL
while still allowing the TM system to
run transactions in parallel as an optimisation.

@@

The second approach, \emph{multiprocessing}, uses multiple instances
of the interpreter itself and runs them in separate OS processes.
Here we actually get parallelism because we have one GIL per
interpreter, but of course we have the overhead of multiple processes
/ interpreters and also need to exchange data between them explicitly
and expensively.

@@

In order to reference these objects, we need references that are valid
in all threads and automatically point to the private copies. Since
an object's offset inside a segment is the same in all segments, we
can use this offset to reference objects. Because all segments are
copies of each other, this \emph{Segment Offset ($SO$)} points to the
private version of an object in all threads\,/\,segments. To then
translate this $SO$ to a real virtual memory address when used inside a
thread, we need to add the thread's segment start address to the
$SO$. The result of this operation is called a \emph{Linear Address
  (LA)}. This is illustrated in Figure \ref{fig:Segment-Addressing}.

@@

In order to eliminate the prohibitive memory requirements of keeping
around $N$ segment copies, we share memory between them. The segments
are initially allocated in a single range of virtual memory by a call
to \lstinline!mmap()!.  As illustrated in Figure
\ref{fig:mmap()-Page-Mapping}, \lstinline!mmap()! creates a mapping
between a range of virtual memory pages and virtual file pages. The
virtual file pages are then mapped lazily by the kernel to real
physical memory pages. The mapping generated by \lstinline!mmap()! is
initially linear but can be changed arbitrarily. Especially, we can
remap so that multiple virtual memory pages map to a single virtual
file page. This is what we use to share memory between the segments
since then we also only require one page of physical memory.

@@

As illustrated in Figure \ref{fig:Page-Remapping}, in our initial
configuration (I) all segments are backed by their own range of
virtual file pages. This is the share-nothing configuration.

We then designate segment 0 to be the \emph{Sharing-Segment}. No
thread gets this segment assigned to it, it simply holds the pages
shared between all threads. So in (II), we remap all virtual pages of
the segments $>0$ to the file pages of our sharing-segment. This is
the fully-shared configuration.

During runtime, we can then privatise single pages in segments $>0$
again by remapping single pages as seen in (III).

Looking back at address translation for object references, we see now
that this is actually a two-step process. First, $\%gs{::}SO$ gets
translated to different linear addresses in different threads by the
CPU. Then, depending on the current mapping of virtual pages to file
pages, these LAs can map to a single file page in the sharing-segment,
or to privatised file pages in the corresponding segments. This
mapping is also performed efficiently by the CPU and can easily be
done on every access to an object.

In summary, $\%gs{::}SO$ is translated efficiently by the CPU to
either a physical memory location which is shared between several
threads/segments, or to a location in memory private to the
segment/thread.  This makes the memory segmentation model for
isolation memory efficient again.

@@

    \protect\caption{Page Remapping: (I) after \texttt{mmap()}. (II) remap all pages to
      segment 0, fully shared memory configuration. (III) privatise single
      pages.\label{fig:Page-Remapping}}
\end{figure}

@@

To detect when to privatise pages, we use write barriers before every
write. When the barrier detects that the object is not in a private
page (or any pages that belong to the object), we remap and copy the
pages to the thread's segment. From now on, the translation of
$\%gs{::}SO$ in this particular segment will resolve to the private
version of the object. Note, the $SO$ used to reference the object does
not change during that process.

@@

The job of barriers is to ensure complete isolation between transactions
and to register the objects in the read or write set. We insert read
and write barriers before reading or modifying an object except if
we statically know an object to be readable or writable already.
\begin{description}
\item [{Read~Barrier:}] Adds the object to the read set of the current
  transaction. Since our two-step address translation automatically
  resolves the reference to the private version of the object on every
  access anyway, the read barrier does not need to do address translation anymore.
\item [{Write~Barrier:}] Adds the object to the read and write set of
  the current transaction and checks if all pages of the object are
  private, doing COW otherwise.\\
  Furthermore, we currently allow only one transaction modifying an
  object at a time. To ensure this, we acquire a write lock on the object
  and also eagerly check for a write-write conflict at this point. If
  there is a conflict, we do some contention management to decide which
  transaction has to wait or abort. Eagerly detecting this kind of conflict
  is not inherent to our system, future experiments may show that we
  want to lift this restriction.
\end{description}

@@

Our TM system is designed as a library that covers all aspects around
transactions and object management. The library consists of two parts:
(I) It provides a simple interface to starting and committing
transactions, as well as the required read and write barriers. (II) It
also includes a \emph{garbage collector (GC)} that is closely
integrated with the TM part (e.g. it shares the write barrier). The
close integration helps in order to know more about the lifetime of an
object, as will be explained in the following sections.

@@

\lstinline!stm_start_transaction()!  starts a transaction. It requires
two arguments, the first being a thread-local data structure and the
second a buffer for use by \lstinline!setjmp()!.
\lstinline!stm_commit_transaction()!  tries to commit the current
transaction. \lstinline!stm_read()!, \lstinline!stm_write()!  perform
a read or a write barrier on an object and \lstinline!stm_allocate()!
allocates a new object with the specified size (must be a multiple of
16). \lstinline!STM_PUSH_ROOT()!  and \lstinline!STM_POP_ROOT()!  push
and pop objects on the shadow stack~\footnote{A stack for pointers to
  GC objects that allows for precise garbage collection. All objects
  on that stack are never seen as garbage and are thus always kept
  alive.}.  Objects have to be saved using this stack around calls
that may cause a GC cycle to happen, and also while there is no
transaction running. In this simplified API, only
\lstinline!stm_allocate()!  and \lstinline!stm_commit_transaction()!
require saving object references.

The type \lstinline!object_t!  is special as it causes the
compiler~\footnote{Clang 3.5 with some patches to this address-space
  256 feature} to make all accesses through it relative to the $\%gs$
register.  With exceptions, nearly all accesses to objects managed by
the TM system should use this type so that the CPU will translate the
reference to the right version of the object.

@@

On startup, we reserve a big range of virtual memory with a call to
\lstinline!mmap()! and partition this space into $N+1$ segments. We
want to run $N$ threads in parallel while segment 0 is designated as
the \emph{sharing-segment} that is never assigned to a thread.

@@

Garbage collection plays a big role in our TM system. The GC is
generational and has two generations: the \emph{young} and the
\emph{old} generation.

@@

To improve this situation even more, we introduce the concept of
\emph{overflow objects}. If a minor collection needs to occur during a
transaction, we empty the nursery and mark each surviving object in
the old object space with an \lstinline!overflow_number!  globally
unique to the current transaction. That way we can still detect in a
medium-fast path inside barriers that the object still belongs to the
current transaction.

@@

As seen in the API (section~\ref{sub:Application-Programming-Interfac}),
we use a \emph{shadow stack} in order to provide precise garbage
collection.  Any time we call a function that possibly triggers a
collection, we need to save the objects that we need afterwards on the
shadow stack using \lstinline!STM_PUSH_ROOT()!.  That way, they will
not be freed. And in case they were young, we get their new location
in the old object space when getting them back from the stack using
\lstinline!STM_POP_ROOT()!.

@@

This area can be seen as a continuous array of bytes that is indexed
from the start of the segment by an object's reference ($SO$) divided
by 16 (this is where the requirement of objects to be of at least 16
bytes in size comes from). Instead of just setting the byte to
\lstinline!true!  if the corresponding object was read, we set it to a
\lstinline!read_version!  belonging to the transaction, which will be
incremented on each commit.  Thereby, we can avoid resetting the bytes
to \lstinline!false!  on commit and only need to do this every 255
transactions. The whole code for the barrier is easily optimisable for
compilers as well as perfectly predictable for CPUs:

@@

The job of the write barrier is twofold: first, it serves as a write
barrier for the garbage collector and second, it supports
copy-on-write and adds objects to the write set of the transaction.

@@

First comes the \emph{GC part}: In any case, the object will be added
to the list of objects that need tracing in the next minor collection
(\lstinline!to_trace!).  This is necessary in case we write a
reference to it that points to a young object. We then need to trace
it during the next minor collection in order to mark the young object
alive and to update its reference to the new location it gets moved
to. The check for \lstinline!is_overflow_obj()!  tells us if the
object was actually created in this transaction. In that case, we do
not need to execute the following \emph{TM part}.  We especially do
not need to privatise the page since no other transaction knows about
these ``old'' objects.

For TM, we first perform a read barrier on the object. We then try to
acquire its write lock. \lstinline!write_locks!  again is a simple
global array of bytes that is indexed with the $SO$ of the object
divided by 16. If we already own the lock, we are done.  If someone
else owns the lock, we will do a write-write contention management
that will abort either us or the current owner of the object.  If we
succeed in acquiring the lock using an atomic
\lstinline!cmp_and_swap!, we need to add the object to the write set
(a simple list called \lstinline!modified_old_objects!)  and privatise
all pages belonging to it (copy-on-write).

In all cases, we remove the \lstinline!WRITE_BARRIER!  flag from the
object before we return. Thus, we never trigger the slow path again
before we do the next minor collection (also part of a commit) or we
start the next transaction.

@@

Aborting a transaction is rather easy. The first step is to reset the
nursery and all associated data structures. The second step is to go
over all objects in the write set (\lstinline!modified_old_objects!)
and reset any modifications in our private pages by copying from the
sharing-segment. What is left is to use \lstinline!longjmp()!  to jump
back to the location initialised by a \lstinline!setjmp()!  in
\lstinline!stm_start_transaction()!.  Increasing the
\lstinline!read_version!  is also done there.

@@

Committing a transaction needs a bit more work. First, we synchronise
all threads so that the committing one is the only one running and all
the others are waiting in a safe point. We then go through the write
set (\lstinline!modified_old_objects!)  and check the corresponding
\lstinline!read_markers!  in other threads/segments. If we detect a
read-write conflict, we do contention management to either abort us or
the other transaction, or to simply wait a bit (see \ref{subsub:contentionmanagement}).

@@

If we want to synchronise all threads, we can rely on this check being
performed regularly. So what we do is to set the
\lstinline!nursery_end!  to $0$ in all segments that we want to
synchronise. The mentioned check will then fail in those segments and
call the slow path. In \lstinline!allocate_slowpath!  they can simply
check for this condition and enter a safe point.

>>>>

If we start multiple such transactions in multiple threads, the TM
system guarantees that the outcome of running the transactions is
\emph{serialisable}. Meaning, the outcome is equal to some sequential
execution of these transactions. This means that the approach provides the same
semantics as using the GIL
while still allowing the TM system to
run transactions in parallel as an optimisation.
\remi{maybe some more explanation of how exactly TM replaces the GIL}

@@

The second approach, \emph{multiprocessing}, uses multiple instances
of the interpreter itself and runs them in separate OS processes.
Here we actually get parallelism because we have one GIL per
interpreter, but of course we have the overhead of multiple processes~/
interpreters and also need to exchange data between them explicitly
and expensively.

@@

In order to reference these objects, we need references that are valid
in all threads and automatically point to the private copies. Since
an object's offset inside a segment is the same in all segments, we
can use this offset to reference objects. Because all segments are
copies of each other, this \emph{Segment Offset ($SO$)} points to the
private version of an object in all threads~/ segments. To then
translate this $SO$ to a real virtual memory address when used inside a
thread, we need to add the thread's segment start address to the
$SO$. The result of this operation is called a \emph{Linear Address
  (LA)}. This is illustrated in Figure \ref{fig:Segment-Addressing}.

@@

In order to eliminate the prohibitive memory requirements of keeping
around $N$ segment copies, we share memory between them. The segments
are initially allocated in a single range of virtual memory by a call
to \lstinline!mmap()!.  As illustrated in Figure
\ref{fig:mmap()-Page-Mapping}, \lstinline!mmap()! creates a mapping
between a range of virtual memory pages and virtual file pages. The
virtual file pages are then mapped lazily by the kernel to real
physical memory pages.

The mapping generated by \lstinline!mmap()! is initially linear but
can be changed arbitrarily. Especially, we can remap so that multiple
virtual memory pages map to a single virtual file page. This is what
we use to share memory between the segments, since we then only
require one page of physical memory for all of them.

@@

As illustrated in Figure \ref{fig:Page-Remapping}, in our initial
configuration (I) all segments are backed by their own range of
virtual file pages. This is the share-nothing configuration where
all threads have private versions of all objects.

We then designate segment~0 to be the \emph{sharing-segment}. No
thread gets this segment assigned to it, it simply holds the pages
shared between threads. So in step (II), we remap all virtual pages of
the segments~$>0$ to the file pages of our sharing-segment. This is
the fully-shared configuration where no threads have private versions
of any objects.

During runtime, we can then privatise single pages in segments~$>0$
again by remapping single pages as seen in (III). All objects in that
page now have a private version in some thread.

Looking back at address translation for object references, we see now
that this is actually a two-step process. First, $\%gs{::}SO$ gets
translated to different linear addresses in different threads by the
CPU. Then, depending on the current mapping of virtual pages to file
pages, these LAs can map to a single file page in the sharing-segment,
or to privatised file pages in the corresponding segments. This
mapping is also performed efficiently by CPUs that have a Memory
Management Unit (MMU) and can easily be done on every access to an
object.

In summary, $\%gs{::}SO$ is translated efficiently by the CPU to
either a physical memory location which is shared between several
threads~/ segments or to a location in memory private to the segment~/
thread. Page sharing makes the memory segmentation model for isolation
memory efficient again.

@@

    \protect\caption{Page Remapping: (I) after \texttt{mmap()}. (II) remap all pages to
      segment~0, fully shared memory configuration. (III) privatise single
      pages.\label{fig:Page-Remapping}}
\end{figure}

@@

To detect when to privatise pages, we use write barriers before every
write to an object. When the barrier detects that the object is not in
a private page (or any pages that belong to the object), we remap and
copy the pages to the thread's segment. From now on, the translation
of $\%gs{::}SO$ in this particular thread will resolve to a private
version of the object automatically. Note that the $SO$ used to reference
the object does not change during that process.

@@

The job of barriers is to ensure complete isolation between transactions
and to register the objects in the read or write set. We insert read
and write barriers before reading or modifying an object, except if
we statically know an object to be readable or writable already.
\begin{description}
\item [{Read~Barrier:}] Adds the object to the read set of the current
  transaction. Since our two-step address translation automatically
  resolves the reference to the private version of the object on every
  access anyway, the read barrier does not have the job to find the
  private version. This job is fully performed by the CPU so that
  read barriers have very little work to do.
\item [{Write~Barrier:}] Adds the object to the read and write set of
  the current transaction and checks if all pages of the object are
  private, doing COW otherwise.\\
  Furthermore, we currently allow only one transaction modifying an
  object at a time. To ensure this, we acquire a write lock on the object
  and also eagerly check for a write-write conflict at this point. If
  there is a conflict, we do some contention management to decide which
  transaction has to wait or abort.\\
  Eagerly detecting this kind of conflict is not inherent to our
  system, future experiments may show that we want to lift this
  restriction.
\end{description}

@@

Our TM system is designed as a library that covers all aspects around
transactions and object management. It is designed for object-oriented
dynamic language VMs as a replacement for the GIL.

The library consists of two parts: (I) It provides a simple interface
to starting and committing transactions, as well as the required read
and write barriers. (II) It also includes a \emph{garbage collector
(GC)} that is closely integrated with the TM part (e.g. it shares the
write barrier). The close integration helps in order to know more
about the lifetime of an object, as will be explained in the following
sections.

@@

\lstinline!stm_start_transaction()!  starts a transaction. It requires
two arguments, the first being a thread-local data structure and the
second a buffer for use by \lstinline!setjmp()!.
\lstinline!stm_commit_transaction()!  tries to commit the current
transaction. \lstinline!stm_read()!, \lstinline!stm_write()!  perform
a read or a write barrier on an object and \lstinline!stm_allocate()!
allocates a new object with the specified size.
 \lstinline!STM_PUSH_ROOT()!  and \lstinline!STM_POP_ROOT()!  push
and pop objects on the shadow stack~\footnote{A stack for pointers to
  GC objects that allows for precise garbage collection. All objects
  on that stack are never seen as garbage and are thus always kept
  alive.}.  Objects have to be saved using this stack around calls
that may cause a GC cycle to happen, and also while there is no
transaction running. In this simplified API, only
\lstinline!stm_allocate()!  and \lstinline!stm_commit_transaction()!
require saving object references.

The type \lstinline!object_t!  is special as it causes the
compiler~\footnote{Clang 3.5 with some patches to its address-space
 256 feature} to make all accesses through it relative to the $\%gs$
register.  With exceptions, nearly all accesses to objects managed by
the TM system should use this type so that the CPU will translate the
reference to the right version of the object.

@@

On startup, we reserve a big range of virtual memory with a call to
\lstinline!mmap()! and partition this space into $N+1$ segments. We
want to run $N$ threads in parallel while segment~0 is designated as
the \emph{sharing-segment} that is never assigned to a thread.

@@

Garbage collection plays a big role in our TM system. The GC is
generational and has two generations: the \emph{young} and the
\emph{old} generation. It is optimised for dynamic languages with
high allocation rates.

@@

To improve this situation even more, we introduce the concept of
\emph{overflow objects}. If a minor collection needs to occur during a
transaction, we empty the nursery and mark each surviving object in
the old object space with an \lstinline!overflow_number!  globally
unique to the current transaction. That way we can still detect in a
medium-fast path inside barriers that the object still belongs to the
current transaction. \remi{so this is where we mention how the GC-STM
integration is useful. highlight more or move to own section?}

@@

As seen in the API (section~\ref{sub:Application-Programming-Interfac}),
we use a \emph{shadow stack} in order to provide precise garbage
collection.  Any time we call a function that possibly triggers a
collection, we need to save the objects that we need afterwards on the
shadow stack using \lstinline!STM_PUSH_ROOT()!.  That way, they will
not be freed. And in case they were young, we get their new location
in the old object space when getting them back from the stack using
\lstinline!STM_POP_ROOT()!. \remi{cite something which explains
shadowstacks in more detail}

@@

This area can be seen as a continuous array of bytes that is indexed
from the start of the segment by an object's reference ($SO$) divided
by 16 (this requires objects to have a size which is dividable by 16).
Instead of just setting the byte to \lstinline!true!  if the
corresponding object was read, we set it to a \lstinline!read_version!
belonging to the transaction, which will be incremented on each
commit.  Thereby, we can avoid resetting the bytes to
\lstinline!false!  on commit and only need to do this every 255
transactions. The whole code for the barrier is easily optimisable for
compilers as well as perfectly predictable for CPUs:

@@

The job of the write barrier is twofold: first, it serves as a write
barrier for the garbage collector and second, it supports COW and adds
objects to the write set of the transaction.

@@

First comes the \emph{GC part}: In any case, the object will be added
to the list of objects that need tracing in the next minor collection
(\lstinline!to_trace!).  This is necessary in case we write a
reference to it that points to a young object. We then need to trace
it during the next minor collection in order to mark the young object
alive and to update its reference to the new location it gets moved
to.

The check for \lstinline!is_overflow_obj()! looks at the
\lstinline!overflow_number!  and tells us if the object was actually
created in this transaction. In that case, we do not need to execute
the following \emph{TM part}.  We especially do not need to privatise
its pages since no other transaction knows about these overflow
objects. Even if they reside in non-private pages, it is guaranteed
that no other transaction can have a reference to them.

For the \emph{TM part}, we first perform a read barrier on the
object. We then try to acquire its write lock. \lstinline!write_locks!
again is a simple global array of bytes that is indexed with the $SO$
of the object divided by 16. If we already own the lock, we are done.
If someone else owns the lock, we will do a write-write contention
management that will abort either us or the current owner of the
object.  If we succeed in acquiring the lock using an atomic
\lstinline!cmp_and_swap!, we need to add the object to the write set
(a simple list called \lstinline!modified_old_objects!)  and privatise
all pages belonging to it (copy-on-write).

In all cases, we remove the \lstinline!WRITE_BARRIER!  flag from the
object before we return. Thus, we never trigger the slow path again
before we do the next minor collection or we start the next
transaction (we always do a minor collection during a commit).

@@

Aborting a transaction is rather easy. The first step is to reset the
nursery and all associated data structures. The second step is to go
over all objects in the write set (\lstinline!modified_old_objects!)
and reset any modifications in our private pages by copying from the
sharing-segment. What is left is to use \lstinline!longjmp()!  to jump
back to the location initialised by a \lstinline!setjmp()!  in
\lstinline!stm_start_transaction()!.  Increasing the
\lstinline!read_version! for the next transaction is also done there.

@@

Committing a transaction needs a bit more work. First, we synchronise
all threads so that the committing one is the only one running and all
the others are waiting in safe points. We then go through the write
set (\lstinline!modified_old_objects!)  and check the corresponding
\lstinline!read_markers!  in other threads~/ segments. If we detect a
read-write conflict, we do contention management to either abort us or
the other transaction, or to simply wait a bit (see \ref{subsub:contentionmanagement}).

@@

If we want to synchronise all threads, we can rely on this check being
performed regularly. So what we do is to set the
\lstinline!nursery_end!  to some small number in all segments that we
want to synchronise. The mentioned check will then fail in those
segments and call the slow path. In \lstinline!allocate_slowpath!
they can simply check for this condition and enter a safe point.
